---
title: "Exploring Cluster Genome Counts"
author: "Nate Olson"
date: "November 9, 2015"
output: html_document
---

## Objective
Summarize number of 16S copies per genome and how the copies are distributed in clusters for different clustering threshold values.

Initial analysis for full length 16S sequences

```{r}
library(magrittr)
library(stringr)
library(dplyr)
library(tidyr)
library(purrr)
library(readr)
library(ggplot2)
library(knitr)
```

### 16S copies per genome
```{r}
setwd("/Users/nolson/multicopy_taxa_assignment")
## note removing rows with gi value as NA - due to carriage returns????
cluster_df <- read_tsv("refseq_clusters/cluster_genome_count.tsv") %>% filter(!is.na(gi))
```

```{r}
copy_count <- cluster_df %>% separate(cluster, c("clus","threshold","id"),  "_",remove = FALSE) %>% filter(threshold == "1.00") %>% group_by(gi) %>% summarize(copies = n())
```

```{r}
ggplot(copy_count) + geom_bar(aes(x = copies)) + theme_bw()
```


```{r}
quantile(copy_count$copies)
```
75% of the genomes have less than 5 copies

## number of clusters by threshold
```{r}
cluster_df %>% separate(cluster, c("clus","threshold","id"),  "_",remove = FALSE) %>% group_by(threshold) %>% summarize( count = length(unique(id))) %>% 
    ggplot()
```


### Number of genomes with copies assigned to multiple clusters
```{r}
cluster_counts <- cluster_df %>% 
    separate(cluster, c("clus","threshold","id"),  "_",remove = FALSE) %>% 
    group_by(threshold, gi) %>% summarize(cl_count = length(unique(cluster)))
```

Summary plot

```{r}
ggplot(cluster_counts) + geom_point(aes(x = threshold, y = cl_count),
                                    position = position_jitter(width = 0.15),
                                    alpha = 0.25) +
    theme_bw() + 
    labs(x = "Cluster Threshold",y = "Clusters per Genome")
```

Summary table

```{r}
cluster_counts %>% group_by(threshold, cl_count) %>% summarize(count = n()) %>% spread(threshold, count, fill = 0) %>% kable()
```

At the genome level only a fraction of the sequences are assigned to different clusters.

#### Number of genomes per cluster
```{r}
cluster_genome_counts <- cluster_df %>% separate(cluster, c("clus","threshold","id"),  "_",remove = FALSE) %>% group_by(cluster, threshold) %>% summarize(genome_count = length(unique(gi)))
```

```{r}
ggplot(cluster_genome_counts) + geom_jitter(aes(x = threshold, y = genome_count), position = position_jitter(width = 0.15)) + scale_y_log10() + theme_bw() + labs(x = "Cluster Threshold", y = "Genomes Per Cluster")
``` 

Number of clusters with more than one genome

```{r}
multi_gi <- cluster_genome_counts %>% rowwise() %>% mutate(multi_gi = ifelse(genome_count > 1, 1,0)) %>% group_by(threshold) %>% summarise(prop_multi = sum(multi_gi)/n())
```

```{r}
ggplot(multi_gi) + geom_bar(aes(x = threshold, y = prop_multi), stat = "identity")
```

Conclusion: for most threshold settings clusters contain sequences from more than one genome.


### Next steps
