---
title: "Checking all 16S multi-copy for duplicates"
author: "Nate Olson"
date: "November 30, 2015"
output: html_document
---
<!-- Fill in title, author, and date above -->

<!-- Resources for creating Rmarkdown documents
    https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf
    https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf
-->

### Purpose
<!-- Reason for performing the analysis -->

### Apparatus
#### Computing Environment 
<!-- System used to run analysis -->
Nate's NIST laptop

#### Software Versions
```{r warning=FALSE, message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(magrittr)
library(knitr)
library(devtools)
library(sapkotaUtils)
```


#### Source Data
Names of source data files and md5 sums for provenance checking

<!-- Name of input data files 
     Make sure to use full file path names for md5 sums
-->
```{r}
file_dir <- "/Users/nolson/multicopy_taxa_assignment/"
source_dat_files <- c(
   seq_names = paste0(file_dir, "refseq_genomes/seq_names.txt"),
   msa = paste0(file_dir, "infernal_aligned/cmd_line_infernal_1.1.fasta")
)
md5table(source_dat_files) %>% kable()
```


### Methods and Results
<!-- Overview of steps used in analysis.  
* Analysis pipeline and results.
* For each step in the analysis
    * summary of software used
    * objective
    * observations
    * notes
-->
Checking for duplicate file names
```{r}
read_csv(source_dat_files['seq_names'],col_names = "name") %>% group_by(name) %>% summarise(count = n()) %>% filter(count > 1) %>% kable()
```

Manually removed duplicates from `refseq_genomes/all_refseq_16S.fasta`

Modifying file names for RAXML - use esl-reformat instead
biostrings is really slow at printing phylip
```{r}
# library(Biostrings)
# seq_file <- readDNAMultipleAlignment(source_dat_files['msa'])
# rownames(seq_file) <- paste0(1:nrow(seq_file)," ",
#                              rownames(seq_file))
# write.phylip(seq_file,
#              paste0(source_dat_files['msa'],
#                     "_renamed.phylip"))
```


### Notes
* Overall observations and conclusions from the analysis
* Next steps

### Session information
```{r}
devtools::session_info()
```

